---
title: "Johnson Creek Ecoregions"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, message=FALSE}
library(tidyverse)
library(sf)
library(vegan)
library(RColorBrewer)
library(viridisLite)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images-jc/",
                      dev="svglite", dev.args=list(fix_text_size=FALSE),
                      fig.height=3, fig.width=8, out.width="1800px")
```

# iNaturalist observations

Download [research-grade iNaturalist observations](https://doi.org/10.15468/dl.ph88g7) from GBIF. Then crop to the [watershed boundary](https://www.inaturalist.org/places/geometry/154205.kml) and color observations by phylum. Draw the creek line from OpenStreetMap. You can see clusters of observations for Crystal Springs, Foster Floodplain, and Powell Butte.

```{r load_data, results='hide'}
#Johnson Creek watershed boundary from iNaturalist place
watershed <- st_read("data/johnsoncreek-inat-154205.kml")

# Johnson Creek stream line from OpenStreetMap
# library(osmdata)
# opq(bbox = st_bbox(watershed)) %>%
#   add_osm_feature(key = "waterway", value = "stream") %>%
#   osmdata_sf() %>% `$`(osm_lines) %>% filter(name=="Johnson Creek") %>% 
#   st_write(creek, "data/johnsoncreek_osm.kml")
creek <- st_read("data/johnsoncreek_osm.kml")

#Portland area for GBIF download
box <- st_as_sfc("POLYGON((-122.9 45.3,-122 45.3,-122 45.8,-122.9 45.8,-122.9 45.3))") %>% st_set_crs("WGS84")

#GBIF data with iNaturalist observations
dat.all <- read_tsv("data/0023880-240906103802322.csv.gz")
rare_cutoff <- 100 #drop rare species
dat.common <- dat.all %>% 
  filter(coordinateUncertaintyInMeters < 1000) %>% 
  group_by(species) %>% filter(n()>rare_cutoff) %>% drop_na(species) %>% 
  mutate(species=factor(species), species_int = as.integer(species)) %>% as.data.frame()

dat.points <- dat.common %>% st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = "WGS84") %>% 
  st_intersection(watershed)

phyla <- c(plant="Tracheophyta", mollusc="Mollusca", vertebrate="Chordata", basidiomycete="Basidiomycota", ascomycete="Ascomycota", arthropod="Arthropoda")
```

```{r load_gbif}
ggplot() + geom_sf(data=watershed, fill="black", color="black") + 
  geom_sf(data=creek, color="grey40", linewidth=2)+
  geom_sf(data=dat.points, aes(color=fct_recode(fct_rev(phylum), !!!phyla)), size=0.3) + theme_void() + 
  scale_color_brewer("",palette = "Set2", guide=guide_legend(override.aes = list(size=4)))
```

```{r divide}
group_by_cell <- function(points, grid, min_obs=0) {
  #assign points to cells
  dat.grid.cells <- st_intersects(grid, points) 
  grid$n_obs <- lengths(dat.grid.cells) #count points in each cell
  
  #drop cells with few observations
  cells.full <- which(grid$n_obs > min_obs)
  grid.full <- grid[cells.full,]
  
  #tally number of each species in each cell
  dat.grid <- t(sapply(dat.grid.cells, function(x) table(points$species[x])))
  rownames(dat.grid) <- grid$grid_id
  dat.grid.full <- dat.grid[cells.full,]
  
  list(data = dat.grid.full, grid = grid.full)
}
```

# Density of observations

Divide the area into hexagons and color by observation density. Missing cells have less than 20 observations.

```{r hex}
hex_dim <- 18 #number of hexagons on each side
hex.sfc <- dat.points %>% st_make_grid(n=c(hex_dim, hex_dim), square = F)
hex <- hex.sfc %>% st_sf() %>% mutate(grid_id = 1:length(lengths(hex.sfc)))

hex_min_obs <- 20 #minimum number of observations allowed in cell
hex.obj <- group_by_cell(dat.points, hex, min_obs=hex_min_obs)
dat.hex <- hex.obj$data
hex.full <- hex.obj$grid
ggplot() +  geom_sf(data=watershed, fill="black", color="black")+
  geom_sf(data=hex.full, aes(fill=n_obs), linewidth=2, color="grey70") + 
  theme_void() + scale_fill_viridis_c("Obs", option="magma") 
```

# Divide into areas of similar population

The number of observations in the hex grid is too uneven. Instead, try to divide the area into polygons of more even populations. First, sample some random seed observations - dense observations areas should have more sampled points. Then, define the region that is closest to each point, makig a Voronoi diagram. The relative variation in observation number per cell is now half that of the hex grid.

```{r voronoi}
n_seeds <- 70
set.seed(1)
dat.sample <- sample_n(dat.points, n_seeds)
tri <- dat.sample %>% st_geometry() %>% do.call(c, .) %>% st_voronoi() %>% 
  st_collection_extract() %>% st_set_crs("WGS84") %>% st_sf() %>%
  mutate(grid_id = 1:nrow(.)) %>%
  st_intersection(watershed)

tri_min_obs <- 8 #minimum number of observations allowed in cell
tri.obj <- group_by_cell(dat.points, tri, min_obs=tri_min_obs)
dat.tri <- tri.obj$data
tri.full <- tri.obj$grid

ggplot() + geom_sf(data=tri.full, aes(fill=n_obs), linewidth=0.5, color="grey70") + 
  geom_sf(data=creek, color="lightblue", linewidth=0.5)+
  theme_void() + scale_fill_viridis_c("Obs", option="magma", limits=c(0,NA))
```

# Cluster similar communities

Look at the observed community composition in each cell, defined as the relative abundance of each taxon. Assign each cell to a cluster of similar communities using the k-means algorithm. Set the number of clusters and assign each a color.

```{r kmeans}
n_clusters <- 6
pal <- c(brewer.pal(8, "Set2"), brewer.pal(12, "Set3"))[1:n_clusters] %>% set_names(1:n_clusters)

set.seed(1)
tri.km <-  dat.tri %>% decostand("hellinger") %>% kmeans(n_clusters)
ggplot() + geom_sf(data=tri.full, aes(fill=factor(tri.km$cluster)), linewidth=0.5, color="grey40") + 
  geom_sf(data=creek, color="black", linewidth=0.5)+
  theme_void() + scale_fill_manual(values=pal, guide="none")
```

# Arrange clusters along gradient

Try to order the clusters by similarity in a linear gradient. This could potentially capture an urban to forested gradient, or riparian to upland, or what people are interested in for different areas. To do this, run a 1 dimensional NMDS ordination of the centers of each cluster, which orders clusters along the number line. Then map the ordered cluster number to a color scale.

```{r gradient}
tri.km.nmds <- metaMDS(tri.km$centers, k=1, trace=F)
tri.full$cluster <- factor(tri.km$cluster, levels=order(t(tri.km.nmds$points)))
ggplot() + geom_sf(data = tri.full, aes(fill=factor(as.integer(cluster))), linewidth=0.5, color="grey80") + 
  geom_sf(data=creek, color="black", linewidth=0.5)+
  theme_void() + scale_fill_viridis_d("Cluster", option="mako", guide="none")
```

# Species that vary across space

To visualize the similarity between communities in each Voronoi cell, run an NMDS ordination on relative abundances. Each point represents a cell, with cells of similar communities clustering together. The colors for each cluster are the same as above. Common species are labeled near the cells where they are most observed and colored by their taxonomic order. The communities are broadly organized by phylum.

```{r nmds, fig.height=8, fig.width=8}
plot_nmds <- function(data, grid, min_to_show) {
  nmds <- metaMDS(data, trace=F)
  specs <- tibble(species = colnames(data), n = colSums(data), show = n>min_to_show) %>%  
    left_join(dat.points %>% as.data.frame %>% count(species, phylum, name="n_total")) %>% 
    mutate(phylum = factor(phylum))
  phylum_pal <- brewer.pal(nlevels(specs$phylum), "Set2")
  
  par(bg="grey30", mar=c(0,0,0,0))
  plot(nmds, type="n", axes=F, ann=F)
  points(nmds, display="sites", col=mako(n_clusters)[grid$cluster], pch=19)
  text(nmds, display="species", cex=0.7, font=4,
       col=if_else(specs$show, phylum_pal[as.integer(specs$phylum)], alpha("black",0)))
  legend("topleft", rev(names(phyla)), fill=phylum_pal, bty="n", text.col="white")
  par(bg="white")
}
plot_nmds(dat.tri, tri.full, 40)
```


