---
title: "Johnson Creek Ecoregions"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, message=FALSE}
library(tidyverse)
library(sf)
library(vegan)
library(RColorBrewer)
library(viridisLite)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images-jc/",
                      dev="svglite", dev.args=list(fix_text_size=FALSE),
                      fig.height=3, fig.width=8)
```

# Load GBIF data

Download research-grade iNaturalist observations from [GBIF as of 19 September 2024](https://doi.org/10.15468/dl.ph88g7). Then crop to the watershed boundary and show observations by kingdom. You can see clusters for Crystal Springs, Foster Floodplain, and Powell Butte.

```{r load_gbif}
#Portland area
#box <- st_as_sfc("POLYGON((-122.9 45.3,-122 45.3,-122 45.8,-122.9 45.8,-122.9 45.3))") %>% st_set_crs("WGS84")
#Johnson Creek watershed boundary from iNaturalist place
box <- st_read("data/johnsoncreek-inat-154205.kml", quiet=T)

dat.all <- read_tsv("data/0023880-240906103802322.csv.gz")

rare_cutoff <- 100 #drop rare species
dat.common <- dat.all %>% 
  filter(coordinateUncertaintyInMeters < 1000) %>% 
  group_by(species) %>% filter(n()>rare_cutoff) %>% drop_na(species) %>% 
  mutate(species=factor(species), species_int = as.integer(species)) %>% as.data.frame()

dat.points <- dat.common %>% st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = "WGS84") %>% 
  st_intersection(box)
ggplot() + geom_sf(data=box, fill="brown", color="brown") + 
  geom_sf(data=dat.points, aes(color=kingdom), size=0.5) + theme_void() + 
  theme(legend.position = "top", plot.background = element_rect(fill="black"), legend.text=element_text(color="grey")) +
  scale_color_brewer("",palette = "Paired", labels=c(Animalia="animal", Plantae="vegetable", Fungi="fungus"),
                     guide=guide_legend(override.aes = list(size=4)))
```

```{r divide}
group_by_cell <- function(points, grid, min_obs=0) {
  #assign points to cells
  dat.grid.cells <- st_intersects(grid, points) 
  grid$n_obs <- lengths(dat.grid.cells) #count points in each cell
  
  #drop cells with few observations
  cells.full <- which(grid$n_obs > min_obs)
  grid.full <- grid[cells.full,]
  
  #tally number of each species in each cell
  dat.grid <- t(sapply(dat.grid.cells, function(x) table(points$species[x])))
  rownames(dat.grid) <- grid$grid_id
  dat.grid.full <- dat.grid[cells.full,]
  
  list(data = dat.grid.full, grid = grid.full)
}
```

# Draw hexagonal grid

Divide the area into hex cells and color by observation density. Missing cells have less than 20 observations.

```{r hex}
hex_dim <- 18 #number of hexagons on each side
hex.sfc <- dat.points %>% st_make_grid(n=c(hex_dim, hex_dim), square = F)
hex <- hex.sfc %>% st_sf() %>% mutate(grid_id = 1:length(lengths(hex.sfc)))

hex_min_obs <- 20 #minimum number of observations allowed in cell
hex.obj <- group_by_cell(dat.points, hex, min_obs=hex_min_obs)
dat.hex <- hex.obj$data
hex.full <- hex.obj$grid
ggplot() +  geom_sf(data=box, fill="grey20")+
  geom_sf(data=hex.full, aes(fill=n_obs), linewidth=2) + 
  theme_void() + scale_fill_viridis_c("Obs", option="magma") + 
  theme(panel.background = element_rect(fill="black"))
```

# Draw Voronoi diagram

The number of observations in the hex grid is too uneven. Instead, try to divide the area into polygons of more even populations. First, sample some random seed observations - dense observations areas should have more sampled points. Then, define the region that is closest to each point. The relative variation in observation number per cell is now half that of the hex grid.

```{r voronoi}
n_seeds <- 100
set.seed(3)
dat.sample <- sample_n(dat.points, n_seeds)
tri <- dat.sample %>% st_geometry() %>% do.call(c, .) %>% st_voronoi() %>% 
  st_collection_extract() %>% st_set_crs("WGS84") %>% st_sf() %>%
  mutate(grid_id = 1:nrow(.)) %>%
  st_intersection(box)

tri.obj <- group_by_cell(dat.points, tri, min_obs=10)
dat.tri <- tri.obj$data
tri.full <- tri.obj$grid

ggplot(tri.full, aes(fill=n_obs)) + geom_sf(linewidth=0.5) + 
  theme_void() + scale_fill_viridis_c("Obs", option="magma", limits=c(0,NA)) + 
  theme(panel.background = element_rect(fill="black"))
```

# k-Means clustering

Look at the observed community composition in each cell, defined as the relative abundance of each taxon. Assign each cell to a cluster of similar communities using the k-means algorithm. The number of clusters can be set. Each cluster is assigned a color.

```{r kmeans}
n_clusters <- 8
pal <- c(brewer.pal(8, "Set2"), brewer.pal(12, "Set3"))[1:n_clusters] %>% set_names(1:n_clusters)

# set.seed(1)
# hex.km <- dat.hex %>% decostand("hellinger") %>% kmeans(n_clusters)
# ggplot(hex.full, aes(fill=factor(hex.km$cluster))) + geom_sf(linewidth=1) + 
#   theme_void() + scale_fill_manual(values=pal, guide="none")

set.seed(2)
tri.km <-  dat.tri %>% decostand("hellinger") %>% kmeans(n_clusters)
ggplot(tri.full, aes(fill=factor(tri.km$cluster))) + geom_sf(linewidth=0.5) + 
  theme_void() + scale_fill_manual(values=pal, guide="none")
```

# Arrange clusters in gradient

Try to order the clusters by similarity in a linear gradient. This could potentially capture an urban to forested gradient, or riparian to upland, or what people are interested in for different areas. To do this, run a 1 dimensional NMDS ordination of the centers of each cluster, which orders clusters along the number line. Then map the ordered cluster number to a color scale.

```{r gradient}
# hex.km.nmds <- metaMDS(hex.km$centers, k=1, trace=F)
# hex.full$cluster <- factor(hex.km$cluster, levels=order(-t(hex.km.nmds$points)))
# ggplot(hex.full, aes(fill=as.integer(cluster))) + geom_sf(linewidth=1) + 
#   theme_void() + scale_fill_viridis_c("Cluster", option="mako")

tri.km.nmds <- metaMDS(tri.km$centers, k=1, trace=F)
tri.full$cluster <- factor(tri.km$cluster, levels=order(t(tri.km.nmds$points)))
ggplot(tri.full, aes(fill=as.integer(cluster))) + geom_sf(linewidth=0.5) + 
  theme_void() + scale_fill_viridis_c("Cluster", option="mako")
```

# NMDS ordination

To visualize the similarity between communities in each voronoi cell, run an NMDS ordination. Each point represents a cell, with cell of similar communities clustering together. The colors for each cluster are the same as above. Selected species are positioned near the cells where they are most common.

```{r nmds, fig.height=8, fig.width=8}
plot_nmds <- function(data, grid) {
  nmds <- metaMDS(data, trace=F)
  par(bg="grey20", mar=c(0,0,0,0))
  plot(nmds, type="n", axes=F, ann=F)
  species.show <- colSums(data)>mean(colSums(data)*1.5)
  points(nmds, display="sites", col=mako(n_clusters)[grid$cluster], pch=19)
  text(nmds, display="species", cex=0.6, col=ifelse(species.show, "grey60", alpha("black",0)))
  par(bg="white")
}
# plot_nmds(dat.hex, hex.full)
plot_nmds(dat.tri, tri.full)
```

