---
title: "Never Home Alone Ecoregions of North America"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, message=FALSE}
library(tidyverse)
library(sf)
sf_use_s2(FALSE)
library(vegan)
library(RColorBrewer)
library(viridisLite)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images-nhana/",
                      dev="svglite", dev.args=list(fix_text_size=FALSE),
                      fig.height=3, fig.width=8, out.width="1800px")
```

# iNaturalist observations

Download iNaturalist observations from [Never Home Alone](https://www.inaturalist.org/projects/never-home-alone-the-wild-life-of-homes). Then color observations by group.

```{r load_data, results='hide'}
ortho <- "+proj=ortho +lat_0=12 +lon_0=-80" #orthographic projection from this vantage point

#North America
world <- rnaturalearth::ne_countries(scale = "small", returnclass = "sf") %>% filter(continent=="North America") %>% st_union()

#iNaturalist observations
phyla <- c(amphibian="Amphibia", `other animal`="Animalia", spider="Arachnida", bird="Aves", chromista="Chromista", fungus="Fungi", insect="Insecta", mammal="Mammalia", plant="Plantae", mollusc="Mollusca", protozoa="Protozoa", reptile="Reptilia")

rare_cutoff <- 20 #drop rare species

dat.points <- read_csv("data/observations-524273.csv.gz", guess_max=1e6) %>% 
  rename(species=scientific_name, decimalLongitude=longitude, decimalLatitude=latitude, phylum=iconic_taxon_name) %>% #GBIF names
  mutate(phylum = phylum %>% fct_infreq() %>% fct_recode(!!!phyla) %>% fct_lump_n(7, other_level="other")) %>% 
  drop_na(decimalLongitude, phylum) %>% 
  as.data.frame() %>% 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = "WGS84") %>% 
  st_intersection(world) %>% 
  group_by(species) %>% filter(n()>rare_cutoff) %>% ungroup() %>% drop_na(species) %>% 
  mutate(species=factor(species), species_int = as.integer(species))

common_names <- dat.points %>% st_drop_geometry() %>% count(species, common_name) %>% drop_na(species) %>% mutate(eithername = coalesce(common_name, species)) %>% select(species, eithername) %>% deframe()
```

```{r load_gbif}
ggplot() + 
  geom_sf(data=world %>% st_transform(ortho), fill="grey90", color="grey90")+
  geom_sf(data=dat.points %>% st_transform(ortho), aes(color=phylum), size=0.3) + theme_void() + 
  scale_color_brewer("",palette = "Set2", guide=guide_legend(override.aes = list(size=4))) 
```

```{r divide}
group_by_cell <- function(points, grid, min_obs=0) {
  #assign points to cells
  dat.grid.cells <- st_intersects(grid, points) 
  grid$n_obs <- lengths(dat.grid.cells) #count points in each cell
  
  #drop cells with few observations
  cells.full <- which(grid$n_obs > min_obs)
  grid.full <- grid[cells.full,]
  
  #tally number of each species in each cell
  dat.grid <- t(sapply(dat.grid.cells, function(x) table(points$species[x])))
  rownames(dat.grid) <- grid$grid_id
  dat.grid.full <- dat.grid[cells.full,]
  
  list(data = dat.grid.full, grid = grid.full)
}
```

# Divide into areas of similar population

The number of observations in the hex grid is too uneven. Instead, try to divide the area into polygons of more even populations. First, sample some random seed observations - dense observations areas should have more sampled points. Then, define the region that is closest to each point, making a Voronoi diagram. The relative variation in observation number per cell is now half that of the hex grid.

```{r voronoi}
n_seeds <- 100
set.seed(1)
dat.sample <- sample_n(dat.points, n_seeds)
tri <- dat.sample %>% st_geometry() %>% do.call(c, .) %>% st_voronoi() %>% 
  st_collection_extract() %>% st_set_crs("WGS84") %>% st_sf() %>%
  mutate(grid_id = 1:nrow(.)) %>%
  st_intersection(world)

tri_min_obs <- 8 #minimum number of observations allowed in cell
tri.obj <- group_by_cell(dat.points, tri, min_obs=tri_min_obs)
dat.tri <- tri.obj$data
tri.full <- tri.obj$grid

ggplot() + geom_sf(data=tri.full%>% st_transform(ortho), aes(fill=n_obs), linewidth=0.2, color="grey70") +
  theme_void() + scale_fill_viridis_c("Obs", option="magma", limits=c(0,NA))
```

# Cluster similar communities

Look at the observed community composition in each cell, defined as the relative abundance of each taxon. Assign each cell to a cluster of similar communities using the k-means algorithm. Set the number of clusters and assign each a color.

```{r kmeans}
n_clusters <- 12
pal <- c(brewer.pal(7, "Set2"), brewer.pal(12, "Set3"))[1:n_clusters] %>% set_names(1:n_clusters)

set.seed(1)
tri.km <-  dat.tri %>% decostand("hellinger") %>% kmeans(n_clusters)
ggplot() + geom_sf(data=tri.full %>% st_transform(ortho), aes(fill=factor(tri.km$cluster)), linewidth=0.2, color="grey40") + 
  theme_void() + scale_fill_manual(values=pal, guide="none")
```

# Species that vary across space

To visualize the similarity between communities in each Voronoi cell, run an NMDS ordination on relative abundances. Each point represents a cell, with cells of similar communities clustering together. The colors for each cluster are the same as above. Common species are labeled near the cells where they are most observed and colored by their taxonomic order. 

```{r nmds, fig.height=8, fig.width=8}
plot_nmds <- function(data, grid, min_to_show) {
  nmds <- metaMDS(data, trace=F)
  specs <- tibble(species = colnames(data), n = colSums(data), show = n>min_to_show) %>%  
    left_join(dat.points %>% as.data.frame %>% count(species, phylum, name="n_total"))
  
  par(bg="white", mar=c(0,0,0,0))
  plot(nmds, type="n", axes=F, ann=F)
  points(nmds, display="sites", col=pal[grid$cluster], pch=19)
  text(nmds, display="species", cex=0.7, font=4,
       label = common_names[colnames(data)],
       col=if_else(specs$show, "grey40", alpha("black",0)))
  par(bg="white")
}
plot_nmds(dat.tri, tri.km, min_to_show=50)
```


